---
title: "Introduction to R vectors"
output: 
---


A vector is a type of object in R that allows us to store data in it. All the data stored in a vector must have the same type. Basic data types are integer, numeric, logical, and character.



**How to create a vector?**


a) Using the combine function: c()

```{r}

# Create a vector "a" storing five numbers

a <- c(3,6,7,10,1)


```




```{r}

# To print the content inside "a", just type in its name

a

```




```{r}

# Add the number 5 to "a"

a <- c(a,5)

```


```{r}

a

```





b) Using the : function

```{r}

b <- 1:5
  
d <- 5:1


b

d

```




c) Using the seq () function

```{r}

e <- seq(from=1, to=5)

e

```




Using the “by” parameter with seq()

```{r}

f<- seq (from= 0, to =10, by=2)

f

```
Help function 

```{r}
?seq
```



What if you want the sequence 10, 8, 6,..., 0? the seg function have more parameters 

```{r}

g<- seq (from=10, to=0, by= -2 )

g

```





d) Using the rep () function

Let’s say you want to create a vector "h" that contains the vector “a” (which we created earlier) repeated three times. 

```{r}

h<- c(a,a,a)

h

```


We can get the same result using the rep () function. 

```{r}

h<- rep(a, 3)

h

```



Let’s say I want to create a vector “k” to contain the vector “f” repeated three times followed by the numbers 10 and 11. How to get it? Think about it for a minute !

```{r}

# ENTER YOUR ANSWER HERE!

k <- c(rep(f,3),10,11)

k
```





Let’s use the parameter “each” with the function rep ()

```{r}
a

```


```{r}

rep (a, each=4)

```






**Indexing a vector**

Indexing refers to the process of retrieving the elements at the different positions of a vector.

Example: Let us create a new vector with positive and negative numbers. Call it vector_y

```{r}

vector_y <- c(23, 5, 67, 3, -2, -8)
vector_y
```



Get first element

```{r}

vector_y[1]

```




Get last element.

To get the last element, it is useful to find out the index of the last element in the vector

```{r}

length(vector_y)

```




```{r}

vector_y[length(vector_y)]

```



```{r}

vector_y [6]

```








Get first 3 elements 

```{r}

vector_y[c(1,2,3)]


# Alternative

vector_y[1:3]

```



Get the first and fifth elements

```{r}

vector_y[c(1,5)]

```



Get all but the second element

```{r}

vector_y

```


```{r}

vector_y[-2]

```




Get all but the second and fifth elements

```{r}

vector_y[-c(2,5)]

```





**Vectorized operations**

Vectorized operations: A function applied to a vector is actually applied individually to each element of that vector.

Many (not all) R built-in functions are vectorized. For example, sqrt (), +, -, *, / … and many others. Some functions are not vectorized since they return a single summary value, for example, sum () and min ().

```{r}
a
```


```{r}
a

a+1

sqrt(a)

a^2

sum(a)

```













