---
title: "Bar Chart using ggplot"
output:
---


```{r}

library (tidyverse)

library(palmerpenguins)

```



```{r}

penguins |> str()

```





**BAR CHART FOR A SINGLE QUALITATIVE VARIABLE**


**Bar chart** to display the **count of penguins for each species**


**Difference between |> and +:**

Use |> to pipe data into a function, whether that function is from dplyr, ggplot2, or base R. 

Use + only to add components together within a ggplot() call. Once you are inside the plot-building expression, everything after ggplot() (e.g., geoms, labs, scales, themes, facets) gets joined with +.



```{r}

penguins |> ggplot(aes(x = species)) + geom_bar() + labs (x = "Species")

```





Sort the previous bar chart by the count (descending order):

```{r}

penguins |> ggplot(aes(x = fct_infreq(species) )) + geom_bar() + labs (x = "Species")

```







Sort the previous bar chart by the count (ascending order):

```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar() + labs (x = "Species")

```







To add a different color for each species:

```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar(aes(color = species)) + labs (x = "Species")

```









```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar(aes(fill = species)) + labs (x = "Species")

```











**BAR CHART FOR THE RELATIONSHIP BETWEEN TWO QUALITATIVE VARIABLES**


**Show the distribution (the frequency) of species (VARIABLE 1) within each island (VARIABLE 2)**

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species)) + labs (x = "Island")

```



```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species), position = "dodge") + labs (x = "Island")

```





To add more ticks to the y axis:

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species), position = "dodge") + labs (x = "Island") + scale_y_continuous(breaks = seq (0, 140, 20))

```






Show the % of penguin species on each island instead of the count:

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar( aes( y = after_stat(count/sum(count))*100 , fill = species), position = "dodge") + labs (x = "Island", y = "Percentage")

```








Show the % of penguins species within each island:

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar( aes( y = after_stat(count/sum(count))*100 , fill = species), position = "fill" ) + labs (x = "Island", y = "Percentage") + scale_y_continuous(labels = scales::percent)

```












**EXERCISES**

The auto data set (i.e., the "mpg" data set) is part of the ggplot2 package.

```{r}

mpg |> str()

```






**Exercise 1**: Create a bar chart showing how many car models each manufacturer has in the mpg dataset. Display the bars in descending order.


WORK INDEPENDENTLY













**Exercise 2**: Create a bar chart showing how many cars have each drive type (drv) within each vehicle class (class).


WORK INDEPENDENTLY


















