---
title: "Bar Chart using ggplot"
output:
---


```{r}

# Then restart R and run:
library(dplyr)
library(tidyverse)
library(palmerpenguins)
library(conflicted)

conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

```



```{r}

penguins |> str()

```





**BAR CHART FOR A SINGLE QUALITATIVE VARIABLE**


**Bar chart** to display the **count of penguins for each species**


**Difference between |> and +:**

Use |> to pipe data into a function, whether that function is from dplyr, ggplot2, or base R. 

Use + only to add components together within a ggplot() call. Once you are inside the plot-building expression, everything after ggplot() (e.g., geoms, labs, scales, themes, facets) gets joined with +.



```{r}

penguins |> ggplot(aes(x = species)) + geom_bar() + labs (x = "Species")

```





Sort the previous bar chart by the count (descending order):

```{r}

penguins |> ggplot(aes(x = fct_infreq(species) )) + geom_bar() + labs (x = "Species")

```







Sort the previous bar chart by the count (ascending order):

```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar() + labs (x = "Species")

```







To add a different color for each species:

```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar(aes(color = species)) + labs (x = "Species")

```









```{r}

penguins |> ggplot(aes(x = fct_rev( fct_infreq(species) )  )) + geom_bar(aes(fill = species)) + labs (x = "Species")

```











**BAR CHART FOR THE RELATIONSHIP BETWEEN TWO QUALITATIVE VARIABLES**


**Show the distribution (the frequency) of species (VARIABLE 1) within each island (VARIABLE 2)**

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species)) + labs (x = "Island")

```



```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species), position = "dodge") + labs (x = "Island")

```





To add more ticks to the y axis:

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar(aes(fill = species), position = "dodge") + labs (x = "Island") + scale_y_continuous(breaks = seq (0, 140, 20))

```






Change the previous plot to show the % of penguin species on each island instead of the count, where the percentages are calculated relative to the grand total across all bars.

```{r}

penguins |> ggplot(aes(x = island)) + geom_bar( aes( y = after_stat(count/sum(count))*100 , fill = species), position = "dodge") + labs (x = "Island", y = "Percentage")

```





What if you wanted to show the percentage of penguin species within each island, where the percentages are calculated within each island? That is, using each island as the total.


```{r}

# Alternative 1

# You get proportions, not percentages

penguins |> ggplot(aes(x = island)) + geom_bar( aes(fill = species), position = "fill" ) + labs (x = "Island", y = "Percentage")

penguins |>
  ggplot(aes(x = island)) +
  geom_bar(aes(fill = species), position = "fill") +
  labs(
    x = "Island",
    y = "Percentage"
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.25),
    labels = scales::percent
  )
```





```{r}

# Alternative 2 (best)

# Percentages are displayed

# 'labels = scales::percent' changes the appearance of the axis labels.

penguins |> ggplot(aes(x = island)) + geom_bar( aes(fill = species), position = "fill" ) + labs (x = "Island", y = "Percentage") + scale_y_continuous(labels = scales::percent)

```












**EXERCISES**

The auto data set (i.e., the "mpg" data set) is part of the ggplot2 package.

```{r}

mpg |> str()

```






**Exercise 1**: Create a bar chart showing how many cars models each manufacturer has in the mpg dataset. Display the bars in descending order.


WORK INDEPENDENTLY

```{r}
mpg |> ggplot(aes(x = fct_infreq(manufacturer))) + geom_bar() + labs(x = "Manufacturer", y = "Number of cars")
```












**Exercise 2**: Create a bar chart showing how many cars have each drive type (drv) within each vehicle class (class).


WORK INDEPENDENTLY


```{r}
mpg |> 
  ggplot(aes(x = class)) +
  geom_bar(aes(fill = drv), position = "dodge") +
  labs(x = "Class")

```
















