---
title: "Means across two groups"
output: 
---


```{r}

library(tidyverse)

```




**Exercise 1)** A study examined whether the perception of the quality of service at a five-star hotel in Jamaica differed by the type of guest (domestic vs foreign). Hotel guests were randomly selected and asked to rate 10 service-related items. Each item was rated on a five-point scale (the higher the score, the higher the perceived quality of service). A score for each guest was obtained by adding the ratings s/he gave for all items. The scores for domestic and foreign guests are recorded in the vectors “domestic_scores” and “foreign_scores” (see next code chunks).


```{r}

domestic_scores <- c(49, 48, 48, 39, 42, 41, 44, 46, 45, 44, 46, 47, 43, 45, 46, 40, 48, 43, 44, 46, 45, 44, 49, 43, 42, 47, 39, 47, 48, 48, 49, 43, 47, 43, 45, 39, 45, 46, 44, 48, 46, 44, 44)
domestic_scores

```


```{r}

foreign_scores <- c(42, 42, 49, 42, 44, 46, 43, 44, 44, 44, 41, 44, 42, 42, 48, 40, 45, 43, 44 ,45, 44, 46, 45, 44, 44, 45, 37, 44, 41, 42, 42, 40, 42, 43, 41, 41, 45, 44, 42, 44, 46, 39, 40)
foreign_scores

```



Management suspects that, on average, the service rating given by domestic guests is higher than the service rating given by foreign guests. Let's analyze the data to find out if their suspicion is backed up by the evidence (i.e., by the collected data).

When conducting the corresponding statistical test, you should consider that the variances of the scores provided by both types of guests are the same.

```{text}
Ho means score domestic <= mean score foreign
Ha means score domestic > mean score foreign

varianzes are the same
```







**EXPLORATORY ANALYSIS**

Let's get some exploratory statistics: The mean, median, and standard deviation of scores for each guest type. Round these numbers to 1 decimal place.

Let's do an exploratory plot: A box plot of scores by guest type.



NOTE: Both samples have the same size (43 observations). See next:

```{r}

length(domestic_scores)

length(foreign_scores)

```


Therefore, in order to be able to use tidyverse functions to do the exploratory analysis, it is better to create a data frame with these two columns. See next:

```{r}

hotel_scores_df <- data.frame( score = c(domestic_scores, foreign_scores), guest_type = c(rep("Domestic", length(domestic_scores)), rep("Foreign", length(foreign_scores))))

```


```{r}

hotel_scores_df |> str()

```





Obtain mean, median, and standard deviation of scores for each guest type. Round these numbers to 1 decimal place.

WORK ON THIS INDEPENDENTLY !!!

```{r}
hotel_scores_df |>
  group_by(guest_type) |>
  summarise(
    mean = round(mean(score), 1),
    median = round(median(score), 1),
    sd = round(sd(score), 1)
  )
```






Do a box plot of scores by guest type.


WORK ON THIS INDEPENDENTLY !!!


```{r}
hotel_scores_df |>
  ggplot(aes(x = guest_type, y = score)) +
  geom_boxplot() +
  labs(title = "Scores", x = "guest type", y = "score")
  
```











**Conclusion from the exploratory analysis**:

The results from the exploratory analysis seem to support the hotel management suspicion because the average and median scores are higher for domestic guests. The box plot confirms that the whole distribution for domestic guests is higher than for foreign ones.

Despite the results from the exploratory analysis, **sometimes** we are required to show statistical evidence for a result obtained from an exploratory analysis. 

That's one of the main goal of this course. To show you how to run statistical tests in certain situations (i.e., for specific types of problems) in case you are required to (or simply decide to) evaluate the statistical significance of exploratory results.

An intuitive way to think about the assessment of the the statistical significance of the results is that it is a way to **find stronger evidence** that our conclusion is valid. 

Since we are trying to compare two samples with respect to a quantitative metric (i.e., the rating score), one relevant test you can apply to assess the statistical significance of the exploratory results is **the two samples t test**.


STOP WORKING ON THIS NOTEBOOK AND LET'S CHECK OUT A PRESENTATION ABOUT THE TWO SAMPLE T TEST.

WE WILL RESUME THIS NOTEBOOK LATER!






**STATISTICAL ASSESSMENT ANALYSIS**

Let's apply the two samples t-test to assess the statistical significance of the results in the hotel guests problem.


**Set up the hypotheses**

Version 1:

Ho: μd <= μf

Ha: μd > μf


Version 2:

Ho: Mean rating domestic guests <= Mean rating foreign guests

Ha: Mean rating domestic guests > Mean rating foreign guests


**NOTE**: a common wrong answer is:

Ho: μ1 <= μ2
Ha: μ1 > μ2



**Run a two samples t test in R**

**Comments about the next code cell**:

The order in which you enter the vectors for the two samples MATTERS! The first vector must always correspond to the first population stated in Ho and Ha.

"var.equal = TRUE" because the problem mentions this: "When conducting the corresponding statistical test, you should consider that the variances of the scores provided by both types of guests are the same."


```{r}

t.test(domestic_scores, foreign_scores, alternative = c("greater"), var.equal = TRUE)

# alternative = c("less")
# alternative = c("two. sided")

```




Decision: The p-value  (0.001144) is less than the established significance level (alpha= 0.05); therefore, the data provide evidence to reject Ho and support Ha. 

Conclusion: The data is backing up the suspicion (claim) that domestic guests give higher average scores than foreign guests.





**EXTRA QUESTION FOR STATISTICAL ASSESSMENT ANALYSIS**


**What kind of error COULD you be making based on the decision you made after running the t test?**


We could be making a type I error. 

Why? Because we rejected Ho. When one rejects Ho after conducting a hypothesis test, IF Ho actually happens to be true (which in reality we do NOT know), we would be making an error.

In this specific problem, a type I error can be described as follows: 

Concluding that domestic guests give higher average scores than foreign guests, WHEN IN REALITY, domestic guests give an average score equal to or lower than that of foreign guests.

Do we have any indication that the risk of making a type I error in this exercise is low? Thankfully yes!

In a statistical test, the probability of making a type 1 error is alpha. Alpha is 0.05 in this problem. Therefore, there is a small chance that we are making such an error in this problem.


**END OF EXTRA QUESTION FOR STATISTICAL ASSESSMENT ANALYSIS**




**Estimate the effect size in this problem**

ANSWER THIS QUESTION LATER. FINISH EXERCISE 2 FIRST. 

YOU NEED TO LEARN ABOUT "practical significance" first before attempting this. You will learn about practical significance by:

a) Reviewing slides from 18 to 22.

b) Watching a YouTube I prepared about this topic (share link to video).














**Exercise 2)** A Starbucks manager suspects that the coffee at two of her stores (stores A and B) yield different caffeine content. To test whether this is true, she takes a random sample of large coffees from both stores and measures the amount of caffeine content in each coffee cup (in mg). The amount of mg of caffeine in a cup of coffee is a random variable that follows a Normal distribution. The results are stored in the vectors “coffeeA” and coffeeB”. 

Run the following code cell to create these two vectors:

```{r}

coffeeA <- c(166, 166, 166, 162, 163, 163, 164, 165, 164, 164, 165, 165, 164, 164, 165, 162, 165, 164, 164, 165)

coffeeB <- c(165, 164, 166, 164, 164, 165, 162, 165, 166, 166, 166, 164, 166, 164, 165, 163, 165, 165, 164, 166)

```



Let's analyze the data to find out if the manager's concern is backed up by the evidence (i.e., by the collected data).


**EXPLORATORY ANALYSIS**

Let's get some exploratory statistics: The mean, median, and standard deviation of the caffeine content for each sample. Round these numbers to 1 decimal place.

Let's do an exploratory plot: A box plot AND a dot plot of caffeine content by store.



NOTE: Both samples have the same size. See next:

```{r}

length(coffeeA)

length(coffeeB)

```


Therefore, in order to be able to use tidyverse functions to do the exploratory analysis, it is better to create a data frame with these two columns. See next:

```{r}

coffee_stores_df <- data.frame( caffeine_content = c(coffeeA, coffeeB), store = c(rep("A", length(coffeeA)), rep("B", length(coffeeB))))

```


```{r}

coffee_stores_df |> str()

```





Obtain mean, median, and standard deviation of the caffeine content for each sample. Round these numbers to 1 decimal place.

WORK ON THIS INDEPENDENTLY !!!




Do a box plot AND a dot plot of caffeine content by store.


WORK ON THIS INDEPENDENTLY !!!









**Conclusion from the exploratory analysis**:

The results from the exploratory analysis do not seem to suggest that the manager's concern is warranted. Both the summary stats and the plots show that the caffeine content levels are similar in both stores. 

In any case, to have a stronger argument to counteract the manager's suspicion, let's run a statistical test to assess the sample results statistical significance.

Since we are trying to compare two samples with respect to a quantitative metric (i.e., the caffeine content), one relevant test you can apply to assess the statistical significance of the exploratory results is **the two samples t test**.





**STATISTICAL ASSESSMENT ANALYSIS**

Let's apply the two samples t-test to assess the statistical significance of the results.

WORK ON THIS INDEPENDENTLY !!!


**Setting up the hypotheses**


COMPLETE IT



**Run a two samples t test**


COMPLETE IT




**What kind of error COULD you be making based on the decision you made after running the t test?**


COMPLETE IT








